---
title: 'Brief Introduction to R package gact'
author: "Palle Duun Rohde, Izel Fourie Sørensen, & Peter Sørensen"
date: "`r Sys.Date()`"
bibliography: qg2021.bib
biblio-style: apalike
link-citations: yes
output:
  bookdown::pdf_document2:
    dev: png
    includes:
      in_header: preamble.tex
  html_document:
    includes:
      in_header: mathjax_header.html
---


```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE, tidy.opts = list(width.cutoff = 70), tidy = TRUE)
eval <- TRUE
echo_solution <- TRUE
message <- FALSE
```

The practical is based on the R package `gact` (Rohde et al.2022)). This package provides an infrastructure for working with large-scale genomic association data linked to different types of genomic features.

The most recent version of `gact` can be obtained from github:

```{r,  eval=FALSE, echo=TRUE}
library(devtools)
devtools::install_github("psoerensen/gact")
```

## Load packages used {.unlisted .unnumbered} 

```{r,  eval=eval, echo=TRUE}
library(gact)
library(qgg)
library(corrplot)
```

## Download and install GDT database {.unlisted .unnumbered} 
The function `gact()` dowload and install the GDT database: 
```{r,  eval=TRUE, echo=TRUE}
# Set working for database
setwd("C://Users//au223366//Dropbox//GitHub//gdtdb")

# Download data bases from repository
GAlist <- gact(version="t2dm-gact-0.0.1", task="download")
```

## Extract data from GDT database {.unlisted .unnumbered} 
The function `getStat()` extract data from the database: 
```{r,  eval=TRUE, echo=TRUE}

# Extract data from T2D for genomic feature Markers
#stat <- getStat(GAlist=GAlist, trait="t2d", feature="Markers")
#head(stat)

# Extract data from T2D for genomic feature Genes
stat <- getStat(GAlist=GAlist, trait="t2d", feature="Genes")
head(stat)
plot(x=stat$stat/stat$m,y=-log10(stat$p), ylab="-log10P", xlab="Tmean",frame.plot=FALSE,main="Genes")

# Extract data from T2D for genomic feature Gene Ontology (GO) where output format is a data frame 
stat <- getStat(GAlist=GAlist, trait="t2d", feature="GO")
head(stat)
plot(x=stat$stat/stat$m,y=-log10(stat$p), ylab="-log10P", xlab="Tmean",frame.plot=FALSE,main="GO")

# Extract data from T2D for genomic feature Gene Ontology (GO) where output format is list 
stat <- getStat(GAlist=GAlist, trait="t2d", feature="GO", format="list", cls=c("p=1e.04","p=0.001","p=0.05","p=0.1","p=0.95"))
str(stat)

# Plot results for genomic feature Gene Ontology (GO) 
colbar <- colorRampPalette(c("#FFFFFF", "#D1E5F0", "#92C5DE",
                             "#4393C3", "#2166AC", "#053061"))
corrplot(-log10(stat$p[1:10,]), is.corr=FALSE,
         tl.cex=0.7, tl.srt=45, col=colbar(6),
         cl.pos="n", mar = c(1, 1, 1, 1))
```

## Extract and write data from GDT database {.unlisted .unnumbered} 
The function `writeStat()` extract and write data from the database: 
```{r,  eval=TRUE, echo=TRUE}
writeStat(GAlist=GAlist, feature="GO", trait="t2d", file.csv="go_t2dm_gcta.csv")
writeStat(GAlist=GAlist, feature="Pathways", trait="t2d", file.csv="pathways_t2dm_gcta.csv")
writeStat(GAlist=GAlist, feature="ProteinComplexes", trait="t2d", file.csv="proteincomplexes_t2dm_gcta.csv")
writeStat(GAlist=GAlist, feature="ChemicalComplexes", trait="t2d", file.csv="chemicalcomplexes_t2dm_gcta.csv")
writeStat(GAlist=GAlist, feature="Genes", trait="t2d", file.csv="genes_t2dm_gcta.csv")
```
